---
title: "Regressions Day 2"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = FALSE)
```

## Load Packages

```{r}
library(broom)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(stargazer)
```

## Recap Last Week

Let's read in the data and clean it:

```{r pressure, echo=FALSE}
ipums_data <- 
  read.csv("acs_2015_data.csv", stringsAsFactors = FALSE)%>%
  mutate(new_educ = case_when(EDUC == 1 ~ 5,
                              EDUC == 2 ~ 9,
                              EDUC == 3 ~ 10,
                              EDUC == 4 ~ 11,
                              EDUC == 5 ~ 12,
                              EDUC == 6 ~ 13,
                              EDUC == 7 ~ 14,
                              EDUC == 8 ~ 15,
                              EDUC == 9 ~ 16, 
                              EDUC == 10 ~ 17, 
                              EDUC == 11 ~ 18), 
         new_sex  = ifelse(SEX == 1, "Male", "Female"),
         new_race = case_when(RACE == 1 ~ "White",
                              RACE == 2 ~ "Black", 
                              RACE == 3 ~ "American Indian or Alaska Native", 
                              RACE %in% c(4,5,6) ~ "Asian or Pacific Islander",
                              RACE == 7 ~ "Other"),
         new_hispan = ifelse(HISPAN %in% c(1,2,3,4), "Hispanic", "Not Hispanic")) %>%
         filter(!is.na(new_educ),
                AGE >= 18,
                AGE <= 65,
                EMPSTAT %in% c(1,2),
                OCC != 0,
                INCWAGE > 0,
                INCWAGE < 999999)
```

Let's run some models

```{r}
baseline_model <- lm(INCWAGE ~ new_educ, data = ipums_data, weights = PERWT)

summary(baseline_model)
```
```{r}
better_model <- lm(INCWAGE ~ new_educ + new_sex, data = ipums_data, weights = PERWT)

summary(better_model)
```

Now come up with a better model:

```{r, eval = FALSE}

```

Now we can output them using stargazer

```{r,results='asis'}
stargazer(baseline_model, better_model, header = FALSE)
```

## Card & Kruger

Read in the data:

```{r}
pa_nj_ff <- read.csv("pa_nj_ff.csv", stringsAsFactors = FALSE)
```

Do some cleaning:

```{r}
pa_nj_ff[pa_nj_ff == "."] <- NA
```


```{r}
pa_nj_cleaned <-
  pa_nj_ff %>%
  mutate(state = ifelse(STATE == 1, "NJ", "PA"),
         location = case_when(SOUTHJ == 1 ~ "Southern NJ",
                              CENTRALJ == 1 ~ "Central NJ",
                              NORTHJ == 1 ~ "Northern NJ",
                              PA1 == 1 ~ "Northeast Suburbs of Philadelphia",
                              PA2 == 1 ~ "Easton",
                              SHORE == 1 ~ "NJ Shore"),
          chain = case_when(CHAIN == 1 ~ "Burger King",
                            CHAIN == 2 ~ "KFC",
                            CHAIN == 3 ~ "Roys",
                            CHAIN == 4 ~ "Wendys"),
          percent_effected_new_min = as.numeric(PCTAFF),
          num_full_emp_1 = as.numeric(EMPFT),
          num_part_emp_1 = as.numeric(EMPPT),
          num_managers_1 = as.numeric(NMGRS),
          starting_wage_1 = as.numeric(WAGE_ST),
          num_hours_1 = as.numeric(HRSOPEN),
          price_soda_1 = as.numeric(PSODE),
          price_fries_1 = as.numeric(PFRY),
          price_entree_1 = as.numeric(PENTREE),
          num_full_emp_2 = as.numeric(EMPFT2),
          num_part_emp_2 = as.numeric(EMPPT2),
          num_managers_2 = as.numeric(NMGRS.1),
          starting_wage_2 = as.numeric(WAGE_ST2),
          num_hours_2 = as.numeric(HRSOPEN2),
          price_soda_2 = as.numeric(PSODE2),
          price_fries_2 = as.numeric(PFRY2),
          price_entree_2 = as.numeric(PENTREE2),
          status_2_interview = case_when(STATUS.2 == 0 ~ "refused to answer",
                                         STATUS.2 == 1 ~ "answered",
                                         STATUS.2 == 2 ~ "closed-renovations",
                                         STATUS.2 == 3 ~ "closed-permanently",
                                         STATUS.2 == 4 ~ "closed-highway",
                                         STATUS.2 == 5 ~ "closed-fire")) %>%
  select(state, location, percent_effected_new_min, num_full_emp_1, num_part_emp_1,
         num_managers_1, starting_wage_1, num_hours_1, price_soda_1, price_fries_1,
         price_entree_1, num_full_emp_2, num_part_emp_2, num_managers_2, starting_wage_2,
         num_hours_2, price_soda_2, price_fries_2, price_entree_2, status_2_interview)

glimpse(pa_nj_cleaned)
```

What is the average wage by state, by fast food chain, before and after the policy goes into effect?

```{r, eval = FALSE}
avg_emp <- pa_nj_cleaned %>%
           group_by() %>%
           summarise()
  
  
avg_emp
```

What is the total number of employees by state, by fast food chain, before and after the policy goes into effect?

```{r}

```

How many stores are not in the second round? please filter them out.

```{r,eval = FALSE}
length(pa_nj_cleaned[pa_nj_cleaned$status_2_interview !=  ] )

pa_nj_cleaned_filered <- pa_nj_cleaned %>%
                         filter(status_2_interview == "")
```

Make a histogram plot of wages in the two periods and use color to indicate states.

```{r, eval = FALSE}
ggplot(pa_nj_cleaned_filered) +
geom_histogram(aes(x = , fill = "Pre-period"), alpha = 0.5) +
geom_histogram(aes(x = , fill = "Post-period"), alpha = 0.5) +
labs(x = "Starting Wage",
     y = "Density",
     fill = "Starting Wage") +
theme_minimal()
```


Make a histogram of employment by region before the minimum wage changes and after.

```{r, eval = FALSE}
ggplot(pa_nj_cleaned_filered) +
geom_histogram(aes(x = , fill = ), alpha = 0.7) +
labs(x = "Number of Full Time Employees",
     y = "Density",
     fill = "Region") +
theme_minimal()
```

Now for some regessions. We need to fix up the data a bit first:

```{r, eval = FALSE}
regression_data <- 
  pa_nj_cleaned_filered %>%
  mutate(change_in_employment_fulltime = num_full_emp_2 - num_full_emp_1,
         change_in_employment_parttime = num_part_emp_2 - num_part_emp_2)

card_krug_full <- lm(change_in_employment_fulltime ~ )
```

And now make them look nice!

```{r}

```

## Logit Regressions

Read in data:

```{r}
grad_school <- read.csv("grad_school.csv", stringsAsFactors = FALSE)
```

Turn rank into a factor variable:

```{r, eval = FALSE}
grad_school_cleaned <- grad_school %>%
                       mutate()
```

Run the logit model:

```{r}
grad_school_logit <- glm(admit ~ gre + gpa + rank, 
                         data = grad_school_cleaned, 
                         family = "binomial")
```

Look at the summary of the output.

```{r}
summary(grad_school_logit)
```

Now use stargazer to make the output look nice:

```{r}

```







