---
title: "Lecture 6 Notes"
author: "Damian Thomas"
date: "2017-10-06"
toc: true
output: pdf_document
---


```{r load-r-packages, include = FALSE}
library(dplyr)
library(ggplot2)
```

\newpage

## Setup

* Etherpad page https://etherpad.net/ecog314_2017-10-06
* Create a directory for today's session 
* Download lecture files https://github.com/wampeh1/Ecog314_Fall2017


## Introduction

* About the instructor
* Announcements


## Objectives

* Write a program and present results using R Markdown
* Define reproduciblity and explain the benefits 
* Review R programming topics
* Illustrate financial literacy concepts 
* List investment strategies for ordinary investors


## Financial Literacy Topics

* Interest rates and inflation
* Risk and diversification
* The relationship between bond prices and interest rates
* The impact that a shorter term can have on total interest payments


## Programming Topics

* Reproducible Research
* R Markdown


## Reproducibile Research


### Reproduciblity

The ability to repeat an analysis and get the same result with the same data

* *What did you do? What did you use?*
* Repeatable



### [Literate Programming](https://en.wikipedia.org/wiki/Literate_programming)

A programming paradigm where programs are written with a mix of natural language explanations of the steps taken and computer code to carry it out. 

* *What were you thinking? Did you explain your decisions?*
* Communicate



### Replicability

The ability to duplicate an analysis and get the same results with new data

* Were your results a fluke? Do your findings hold using different statistical samples?
* Corroborate   



## Introduce R Markdown

R Markdown package:

* Combine R code, output, and formatted text
* Encourages Reproducible Research through Literate Programming.
* Creates nicely formatted documents that include output from R
* Useful for communicating results, collaborating with others, and documenting your work

References 

* http://rstudio.com/cheatsheets
* http://r4ds.had.co.nz/r-markdown.html



## Live coding

* Create an R Markdown Document in RStudio
* Demonstrate R Markdown features


## Create an R Markdown Document in RStudio

1. From the menu: File >> New >> R Markdown
2. Take the defaults
4. Click OK 

This opens a new file in the source pane, with a template created by RStudio.


## Main parts of an R Markdown document

1. YAML header
2. R code chunks
3. Markdown text
    

## Rendering a document

1. Save the file with `.Rmd` file extension
2. Knit the document 
    * Click on the "Knit" button,
    * Type `Ctrl + Shift + K`
    * Use the menu: File >> Knit Document
    * console command
    

## Editing a Document

Refer to http://rstudio.com/cheatsheets

 

### Header

* title
* date
* author
* change the output format
    * `html_document`
    * `pdf_document`
    * `beamer_presentation`



### Markdown text

Add text to the body

* Add / delete text
* bold, italics, underline
* headings
* lists
* links
* tables



### LaTeX (Math)

Reference: https://en.wikibooks.org/wiki/LaTeX/Mathematics

* Add an inline equation: `$ E = mc^{2} $`
* Add a centered Equation: `$$ y = \beta_{0} + \beta_{1}x_{1} + \beta_{2}x_{2} + \epsilon $$` 
* Complex math: $\forall x \in \mathbb{R}^1, \sqrt{x} = i \implies x = -1$


## Code Blocks

### Edit the code block with the summary

* change the data frame to `mtcars`
* run the `str` function
* print the data frame

### Edit the code block with the plot

* load ggplot2 library
* create a scatterplot of speed vs stopping distance

### Run code blocks interactively

* line-by-line
* entire code block



### Code block options

* specify programming language
* code chunk name
* `echo = TRUE` -- Show the code
* `results = "hide"` -- run the code, show the code, but do not show the results
* `include = FALSE` -- run the code, but do not show either the code or the results
* `eval = FALSE` -- show the code, but do not run it
* `cache = TRUE` -- save the output
* `results = "asis"`


## Inline R Code

Enclose the expression in backticks `` `r `` with the name of the language 

`Pick a random number between 1 and 10: ` `` `r `` `sample(1:10, size = 1)` `` ` ``
    
Renders as:

Pick a random number between 1 and 10: `r sample(1:10, size = 1)`


## Start new document

* Create a new R Markdown file
* Take the defaults
* Clear the template

---

## Add an introduction

*What do we intend to do?*

The FINRA Investor Education Foundation conducted a survey to study how people 
manage their resources and make financial decisions. The survey includes
several benchmark questions to assess understanding of key financial concepts.
In today's class we will try a few of these benchmark questions, 
illustrate the concepts using R code, and explore the nationwide survey results.

* Insert a link to the FINRA homepage: http://www.usfinancialcapability.org/
* Save the file, knit it, and click on the link in the rendered document
    
## Take the quiz

http://www.usfinancialcapability.org/quiz.php


## Take notes

Describe what we are doing: taking the quiz, getting familiar with the survey.

* Add two sections with headings to the document: 
    1. "Taking the online survey", "About the survey", etc.
    
> *In order to familiarize ourselves with the survey, we took the online financial literacy quiz. The first question deals with interest earning savings...*

    2. "References"

> "Financial Literacy Quiz," Take the Financial Literacy Quiz, FINRA Investor Education Foundation, Accessed October 6, 2017, http://www.usfinancialcapability.org/quiz.php.



## Question 1

Suppose you have \$100 in a savings account earning 2 percent interest a year. After five years, how much would you have?

a. More than \$102 
b. Exactly \$102 
c. Less than \$102 
d. Don't Know 

* Key concept: interest earning assets gain value as time passes


To show that we have answered correctly, we will model the accumulation of interest using R

Given: 

* $r$ is the interest rate where $r \in [0, 1]$
* $b$ is a vector of account balances
* $b_{t} = b_{t-1} + (rb_{t-1})$ 

$$ b = \left[ \begin{matrix} b_{t}  \\ b_{t+1} \\ \vdots \\ b_{t+n} \end{matrix} \right] $$  

Goal: create a vector where each element is the account balance at a different point in time


## Question 1: R code

To implement this mental model in R, start with the fixed parameters, then use a for-loop to compute the interest in each period

```{r question-1-code, eval = TRUE, echo = TRUE}
interest_rate <- 0.02
balance <- 100 # starting balance
n <- 5

for ( i in 1:n ) {
    t <- i + 1
    balance[t] <- balance[t-1]  + (interest_rate * balance[t-1])
}
```

Based on this model we can confirm that: after 5 periods of compounding the account balance will be `r balance[6]`. A complete table of values is presented below.

```{r account-balance-table}
# take the vector and create a data frame

account <- data.frame(
    t = seq(0, n),
    balance = balance
)

knitr::kable(account, caption = "Table of Account Balances")
```

Let's visualize these results.  
* Add a code block 
* load dplyr and ggplot2
* Suppress the messages and hide the code

```{r quiz-question-1-plot, echo = TRUE, fig.align = "center"}
account %>%
    ggplot(mapping = aes(x = t, y = balance)) +
    geom_bar(stat = "identity") +
    geom_hline(yintercept = 102) + 
    labs(title = "Account Balance After 5 years Earning 2% Interest on $100",
         caption = "The horizontal line represents a balance of $102",
         x = "Time period",
         y = "Dollars")
```


## Question 2

Imagine that the interest rate on your savings account is 1 percent a year and inflation is 2 percent a year. After one year, would the money in the account buy more than it does today, exactly the same or less than today?

* Inflation reduces the buying power of the savings account. The nominal value is not the same thing as real value. 

* Let's compute the Inflation adjusted returns. Using the investopedia rule.
* Add a reference to the source at the end of the document


> "Inflation-Adjusted Returns", Inflation-Adjusted Returns, accessed 2017-10-07, http://www.investopedia.com/terms/i/inflation_adjusted_return.asp


## Question 2: R code

```{r quiz-question-2, echo = TRUE}
years <- 2017:2030
starting_balance <- 100
interest_rate  <- 0.01
inflation_rate <- 0.02
adjusted_rate <- ((1 + interest_rate) / (1 + inflation_rate)) - 1
adjusted_rate

nominal_balance  <- c()
nominal_gains    <- c()
adjusted_gains   <- c()
adjusted_balance <- c()

for ( t in seq_along(years) ) {
    if ( t == 1 ) {
        nominal_gains[t] <- 0
        adjusted_gains[t] <- 0
        nominal_balance[t] <- starting_balance
        adjusted_balance[t] <- nominal_balance[t]
    } else {
        nominal_gains[t] <- interest_rate * nominal_balance[t-1]
        adjusted_gains[t] <- adjusted_rate * adjusted_balance[t-1]
        nominal_balance[t] <- nominal_balance[t-1] + nominal_gains[t-1]
        adjusted_balance[t] <- adjusted_balance[t-1] + adjusted_gains[t-1]
    }
}

account <- data.frame(year = years,
                      nominal_gains,
                      adjusted_gains,
                      nominal_balance,
                      adjusted_balance)

account
```


## Question 2: Plots

```{r plot-q2-nominal}
account %>%
    ggplot(mapping = aes(x = year, y = nominal_balance)) +
    geom_bar(stat = "identity") +
    geom_hline(yintercept = 100) + 
    labs(title = paste("Change in Nominal Account Balance"),
         caption = "The horizontal line represents a balance of $100",
         x = "Year",
         y = "Dollars")
```



```{r plot-q2-inflation-adjusted}
account %>%
    ggplot(mapping = aes(x = year, y = adjusted_balance)) +
    geom_bar(stat = "identity") +
    geom_hline(yintercept = 100) + 
    labs(title = paste("Change in Account Balance, Adjusted for inflation"),
         caption = paste("The horizontal line represents the value of $100 in", min(account$year)),
         x = "Year",
         y = "Dollars")
```


## Survey Data

Let's look at the data and see how people responded to these questions in the survey.

## Create a new section heading

Create a section entitled "Data". Add text to describe where we are getting our data

Data for the 2015 National Financial Capability Study (NFCS) 
are available for download from the foundation's website. To 
keep track of what we are doing we will use R to download 
the 2015 State-by-State Respondent-Level Data.
    

## Create directories

Store raw data separately. To maintain reproducibility, do not alter the raw input data.

```{r create-raw-data-directory}
# location for raw source data downloaded from the internet
raw_dir <- file.path("data", "raw")

if ( !dir.exists(raw_dir) ) {
    dir.create(raw_dir)
}
```

## Download the raw data, unzip


```{r download-2015-data, results = "hide"}

# Source data
link <- "http://www.usfinancialcapability.org/downloads/NFCS_2015_State_by_State_Data_Excel.zip"

# Path for a local copy of the file
save_as <- file.path(raw_dir, basename(link))


# Conditionally downlad and unzip the file
if ( !file.exists(save_as) ) { 
    
    # Download the raw data file if we don't already have it
    download.file(link, save_as)
    unzip(save_as, exdir = raw_dir)  

} else if ( !file.exists(file.path(raw_dir, "NFCS 2015 State Data 160619.csv")) ) {
    
    # Extract all raw data files from an existing zip file
    unzip(zip_file, exdir = raw_dir)  
    
} else {

    list.files(raw_dir)

}
```


## Import the raw data

```{r read-raw-csv}
raw_data <- read.csv(file = file.path(raw_dir, "NFCS 2015 State Data 160619.csv"))
```



## Clean the data, filter an reformat variables

The raw data set has `r ncol(raw_data)` variables and we aren't going to use every variable, so we will restrict the data to a few columns and use more descriptive names. The columns we plan to keep are:

```{r variable-list}
raw_variables <- c(
    "NFCSID" = "Respondent ID",
    "STATEQ" = "State ID",
    "A3"     = "Gender",
    "A3Ar_w" = "Age Group",
    "M6"     = "Question 1: Interest rate",
    "M7"     = "Question 2: Interest rates and inflation",
    "M8"     = "Question 3: Interest rates and bond prices",
    "M31"    = "Question 4: Compounding interest",
    "M9"     = "Question 5: Compounding and loan term",
    "M10"    = "Question 6: Diversification",
    "wgt_n2" = "National-level weight",
    "wgt_s3" = "State-level weight"
    )

data.frame(variable = names(raw_variables), 
           description = unname(raw_variables)) %>%
    knitr::kable()
```

Also, we will need to tidy the data by applying labels to the coded values.

```{r value-labels, echo = FALSE}

stateq_label <- c(
     "1" = "Alabama",
     "2" = "Alaska",
     "3" = "Arizona",
     "4" = "Arkansas",
     "5" = "California",
     "6" = "Colorado",
     "7" = "Connecticut",
     "8" = "Delaware",
     "9" = "District of Columbia",
    "10" = "Florida",
    "11" = "Georgia",
    "12" = "Hawaii",
    "13" = "Idaho",
    "14" = "Illinois",
    "15" = "Indiana",
    "16" = "Iowa",
    "17" = "Kansas",
    "18" = "Kentucky",
    "19" = "Louisiana",
    "20" = "Maine",
    "21" = "Maryland",
    "22" = "Massachusetts",
    "23" = "Michigan",
    "24" = "Minnesota",
    "25" = "Mississippi",
    "26" = "Missouri",
    "27" = "Montana",
    "28" = "Nebraska",
    "29" = "Nevada",
    "30" = "New Hampshire",
    "31" = "New Jersey",
    "32" = "New Mexico",
    "33" = "New York",
    "34" = "North Carolina",
    "35" = "North Dakota",
    "36" = "Ohio",
    "37" = "Oklahoma",
    "38" = "Oregon",
    "39" = "Pennsylvania",
    "40" = "Rhode Island",
    "41" = "South Carolina",
    "42" = "South Dakota",
    "43" = "Tennessee",
    "44" = "Texas",
    "45" = "Utah",
    "46" = "Vermont",
    "47" = "Virginia",
    "48" = "Washington",
    "49" = "West Virginia",
    "50" = "Wisconsin",
    "51" = "Wyoming"
)

a3_label <- c(
    "1" = "Male",
    "2" = "Female"
)

a3ar_w_label <- c(
    "1" = "18-24",
    "2" = "25-34",
    "3" = "35-44",
    "4" = "45-54",
    "5" = "55-64",
    "6" = "65+"
)

m6_label <- c(
     "1" = "More than $102",
     "2" = "Exactly $102",
     "3" = "Less than $102",
    "98" = "Don't know",
    "99" = "Prefer not to say"
)
    
m7_label <- c(
     "1" = "More than today",
     "2" = "Exactly the same",
     "3" = "Less than today",
    "98" = "Don't know",
    "99" = "Prefer not to say"
)


m8_label <- c(
     "1" = "They will rise",
     "2" = "They will fall",
     "3" = "They will stay the same",
     "4" = "There is no relationship between bond prices and the interes",
    "98" = "Don't know",
    "99" = "Prefer not to say"
)

m31_label <- c(
     "1" = "Less than 2 years",
     "2" = "At least 2 years but less than 5 years",
     "3" = "At least 5 years but less than 10 years",
     "4" = "At least 10 years",
    "98" = "Don't know",
    "99" = "Prefer not to say"
)

m9_label <- c(
     "1" = "True",
     "2" = "False",
    "98" = "Don't know",
    "99" = "Prefer not to say"
)

m10_label <- c(
     "1" = "True",
     "2" = "False",
    "98" = "Don't know",
    "99" = "Prefer not to say"
)

```



```{r create-tidy-data}
clean_data <- raw_data %>%
    #select_(.dots = names(raw_variables)) %>% # using a character vector to select columns
    mutate(id = NFCSID,
           state = stateq_label[STATEQ],
           #question_1 = factor(M6, levels = names(m6_label), labels = unname(m6_label)),
           question_1 = m6_label[M6],
           question_2 = m7_label[M7],
           question_1_correct = ifelse(question_1 == "More than $102", TRUE, FALSE),
           question_2_correct = ifelse(question_2 == "Less than today", TRUE, FALSE),
           state_weight = wgt_s3,
           national_weight = wgt_n2) %>%
    select(id, state, question_1, question_2, question_1_correct, question_2_correct, state_weight, national_weight)

str(clean_data)
```

Explore data

```{r q1-by-state-table}
clean_data %>%
    filter(!is.na(question_1)) %>%
    group_by(state) %>%
    summarize(q1_correct_pct = 100 * sum(state_weight * question_1_correct) / sum(state_weight)) %>% 
    arrange(desc(q1_correct_pct)) %>%
    knitr::kable()
```

```{r q1-by-state, fig.height = 10}
clean_data %>%
    filter(!is.na(question_1)) %>%
    group_by(state) %>%
    summarize(q1_correct_pct = 100 * sum(state_weight * question_1_correct) / sum(state_weight)) %>% 
    arrange(q1_correct_pct) %>%
    mutate(state = factor(state, state)) %>% 
    ggplot(mapping = aes(x = state, weight = q1_correct_pct)) +
    geom_bar(stat = "count") +
    coord_flip()

```

```{r q2-by-state-table}
clean_data %>%
    filter(!is.na(question_2)) %>%
    group_by(state) %>%
    summarize(q2_correct_pct = 100 * sum(state_weight * question_2_correct) / sum(state_weight)) %>% 
    arrange(desc(q2_correct_pct)) %>%
    knitr::kable()
```


```{r q2-bystate-plot, fig.height = 10 }
clean_data %>%
    filter(!is.na(question_2)) %>%
    group_by(state) %>%
    summarize(q2_correct_pct = 100 * sum(state_weight * question_2_correct) / sum(state_weight)) %>% 
    arrange(q2_correct_pct) %>%
    mutate(state = factor(state, state)) %>% 
    ggplot(mapping = aes(x = state, weight = q2_correct_pct)) +
    geom_bar(stat = "count") +
    coord_flip()
```

```{r histograms-for-each-state, eval = FALSE, echo = FALSE}
clean_data %>% 
    group_by(state, question_1) %>%
    summarise(wt = sum(state_weight)) %>%
    ggplot(mapping = aes(x = question_1, group = state, weight = wt)) +
    geom_bar(stat = "count") +
    facet_wrap( ~ state, ncol = 7)

```


\newpage

## References



"R for Data Science", http://r4ds.had.co.nz/r-markdown.html

"R Markdown Cheatsheet", http://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf

"LaTeX mathematics", https://en.wikibooks.org/wiki/LaTeX/Mathematics

"Literate Programming", https://en.wikipedia.org/wiki/Literate_programming

"Inflation-Adjusted Returns", Inflation-Adjusted Returns, accessed 2017-10-07, http://www.investopedia.com/terms/i/inflation_adjusted_return.asp

Freakonomics Radio Podcast. "Everything You Always Wanted to Know About Money (But Were Afraid to Ask)". Posted August 2, 2017 at 11:00pm. [http://freakonomics.com/podcast/everything-always-wanted-know-money-afraid-ask/]()

Chicago Manual of Style Online. "Notes and Bibliography: Sample Citations." 2017.  [http://www.chicagomanualofstyle.org/tools_citationguide/citation-guide-1.html#cg-website]()

American Economic Association. "Sample References." 2017. [https://www.aeaweb.org/journals/policies/sample-references]()

