---
title: "Communicating Results with R Markdown"
author: "Damian Thomas"
date: "2017-10-06"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Objectives for this session

* Use Markdown syntax to format text
* Use R Markdown features to embed code in a document
* Complete an analysis and present results using R Markdown
* Output multiple document formats using R Markdown (html, pdf, slides)
* List the advantages of R markdown 
* Define reproduciblity and differentiate from related concepts (replication, tidyness, literate programming) 


## What is R Markdown?

*"R Markdown provides an unified authoring framework for data science, combining your code, its results, and your prose commentary. R Markdown documents are fully reproducible and support dozens of output formats, like PDFs, Word files, slideshows, and more."*

[Garrett Grolemund and Hadley Wickham. R for Data Science (Ch 27)]([http://r4ds.had.co.nz/r-markdown.html)


## What is R Markdown?

* R code + Formatted text = Complete record of analysis


## Example: R Markdown Report

*Showcase compelling examples*


## How-To: Create a new R Markdown document

1. From the menu: File >> New >> R Markdown
2. Give the file a name with a `.Rmd` extension
3. Enter a title (optional)
4. Click OK


## R Markdown Document Structure

* Front Matter
* Markdown text
* R code blocks
* inline code


## How-To: Knit an R Markdown document

* Click on the "Knit" button
* Type `Ctrl + Shift + K`
* Use the menu: File >> Knit Document


## Edit a R Markdown Document

* Change the title
* Change text
* Change code block
* Error handling


## What happens in the background when you knit an R Markdown document

* Interpret the markdown text
* Run the code chunks
* insert the results into the document
* output a new file, if all is well


## How-To: Knit different file formats

* Click the dropdown next to the Knit button
    * select output format
    
* Change the the document header
    * `output: html_document`
    * `output: pdf_document`
    * `output: beamer_presentation`


## More edits


* Hide code block
* Skip code block
* Inline Code
* Display plot
* Display table

## Motivation / beneits


* Share your work with others
* Check your work and update it, time saver in the future
* Answer questions about previous work 
* Allow others build on what you have done, instead of working to recreate it
* Apply previously developed process to new data
* Encourage trust in the results, if peers can plausibly recreate
* Capture what you did, how you did it, and what you were thinking in a single presentation-ready document



## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

